Decisions Decisions Decisions: Carrying out Latent Profile Analysis in Accordance With Best Practices Using the tidyLPA R package

### 2020-12-29 paper.rmd
re-run analysis from top-to-bottom

author: 
  - name          : "Joshua Rosenberg"
    affiliation   : "1"
    corresponding : yes
    address       : "1122 Volunteer Blvd., Knoxville, TN, 37996"
    email         : "jmrosenberg@utk.edu"
  - name          : "Caspar van Lissa"
    affiliation   : "2"
  - name          : "Jennifer Schmidt"
    affiliation   : "3"
  - name          : "Patrick Beymer"
    affiliation   : "3"
  - name          : "Daniel Anderson"
    affiliation   : "4"
  - name          : "Matthew Schell"
    affiliation   : "3"

affiliation:
  - id            : "1"
    institution   : "University of Tennessee, Knoxville"
  - id            : "2"
    institution   : "Utrecht University"
  - id            : "3"
    institution   : "Michigan State University"
  - id            : "4"
    institution   : "University of Oregon"
  - id            : "5"

abstract: 
keywords          : "Latent Profile Analysis, mixture models, finite mixture models, tutorial, R, MPlus, mclust"
wordcount         : ""

@@

```{r setup, include = FALSE}
library(papaja)
```

```{r analysis-preferences, echo = FALSE}
# Seed for random number generation
set.seed(42)
library(tidyverse)
library(simstudy)

@@

Students, may (reasonably) ask, though, are grades *really* normally distributed?

```{r, include = FALSE, eval = TRUE}
def <- defData(varname = "x1", formula = 65, variance = 30)
def <- defData(def, varname = "x2", formula = 85, variance = 23)
def <- defData(def, varname = "xMix", 
               formula = "x1 | .5 + x2 | .5", 
               dist = "mixture")

set.seed(2716)
dx <- genData(1000, def)
```

Are their one distribution?

```{r, include = FALSE, eval = TRUE}
ggplot(dx, aes(x = xMix)) +
  geom_density() +
  papaja::theme_apa() +
  ylim(0, .10)
```

Or two?

```{r, include = FALSE, eval = TRUE}
dx %>% 
  gather(key, val, x1:x2) %>% 
  ggplot(aes(x = val, fill = key)) +
  geom_density(alpha = .3) +
  papaja::theme_apa() +
  ylim(0, .10)
```

This kind of distribution (that is *bimodal*) is not exclusive to grades. For
teachers, psychologists, researchers, and friends and family members, people are
highly-complex and not easily able to be characterized by *one* characteristic
or personality trait--and its distribution. In the social sciences, broadly, and
in psychology, particularly, a statistical method that can be used to describe
how people, in their individual particularities, may have similarities on the
basis of some set of measures through which they can be grouped in meaningful,
distinctive ways. This approach, which has a provenance in developmental (or
*person-oriented*) approaches (Bergman & El-Khouri, 1997; Magnusson &
Cairns, 1996; see Linnenbrink-Garcia & Wormington, 2017, for a recent review) is
an example of a general mixture model (Harring & Hodis, 2016; Pastor, Barron,
Miller, & Davis, 2007).

In this tutorial, we aim to describe one of the most commonly-used--and relevant
to psychologists---application of the general mixture model, to cases for which
all of the variables for which (relatively) homogeneous groups are identified
from among a (relatively) heterogeneous sample are continuos, *latent profile
analysis* (LPA). We note that scholars using the approach, especially in
developmental and educational psychology, use cluster analytic methods. Such an
approach has some similarities with LPA (CITE paper showing how k-means is
similar to the simplest model specification of LPA?), though LPA also has a
number of benefits (Linnenbrink-Garcia & Wormington, 2017), and we we focus on
LPA in this paper.

After describing the method and some examples of its use, we provide a tutorial
for carrying out LPA in the context of a freely-available, open-source
statistical software package we created for R (R Core Team, 2019), tidyLPA.
Finally, we offer some ideas about best practices and informed recommendations
for researchers aiming to use LPA in their applied work, and conclude with
reflections on the role of statistical software--especially software that is
freely-available, open-source, and highly-performant--in the psychological
sciences.

# Latent Profile Analysis

The goal of LPA is estimate the parameters for a number of distributions
(typically multivariate) from a single data set. Thus, such an approach is
model-based, and some descriptions in the literature refer to it as model-based
clustering (Hennig, Meila, Murtagh, & Rocci, 2015; Scrucca, Fop, Murphy, &
Raftery, 2017). Thus, one distinction between LPA and other, similar cluster
analytic approaches is that LPA is model-based; instead of using algorithms to
group together cases, LPA seeks to estimate parameters - in terms of variances
and covariances and how they are the same or different across profiles - that
best characterize the different distributions. Then, this approach seeks to
assign to each observation a probability that the observation is a sample from
the population associated with each profile (or mixture component).

@@

We note that challenge facing the analyst using LPA is that these
parameters and the distinct model parameterizations that can be estimated is the
different terminology used. As one example, Scrucca et al. (2017) refer to these
parameterizations not in terms of whether and how parameters are estimated, but
rather in terms of the geometric properties of the distributions that result
from particular parameterizations. Muthen and Muthen (1997-2017) and others
(Pastor et al., 2007) commonly refer to local independence to mean that the
covariances are fixed to zero (also described as the specification of the
covariance matrix as “diagonal,” because only the diagonal components, or the
variances, are estimated).

In general, as more parameters are estimated (i.e., those that are fixed to zero
are estimated as being equal across profiles; or those estimated as being equal
across profiles are freely-estimated across them), the model becomes more
complex; the model may fit better, but also be overfit, meaning that the
profiles identified may be challenging to replicate with another, separate data
set. Even still, flexibility in terms of which models can be estimated also has
affordances. For example, the varying means, equal variances, and covariances
fixed to 0. A researcher might choose this model specification if she wants to
model the variables to be used to create profiles that are independent. This
model is very simple, as no covariances are estimated and the variances are
estimated to be the same across profiles. As we estimate more parameters (and
decrease the degrees of freedom), we are more likely to fit the data, but less
likely to be able to replicate the model with a second set of data. In other
words, more parameters may mean a loss of external validity. As we progress
toward more complex models (with increasingly complex parameterization), then we
are more likely to fit the data better. More information on the model
parameterizations are discussed in the context of the software tool tidyLPA that
we have developed.

# tidyLPA

## Tools other than tidyLPA

@@

In R, a number of tools *can* be used to carry out LPA. OpenMx can be used for
this purpose (and to specify almost any model possible to specify within a
latent variable modeling approach). However, while OpenMx is very flexible, it
can also be challenging to use. Other tools in R allow for estimating Gaussian
mixture models, or models of multivariate Gaussian (or normal) distributions. In
addition to following the same general approach, using tools that are designed
for Gaussian mixture modeling have other benefits, some efficiency-related (see
RMixMod, which uses compiled C++ code) and others in terms of ease-of-use (i.e.,
the plot methods built-in to RMixMod, mclust, and other tools). However, they
also have some drawbacks, in that it can be difficult to translate between the
model specifications, which are often described in terms of the geometric
properties of the multivariate distributions being estimated (i.e., “spherical,
equal volume”), rather than in terms of whether and how the means, variances,
and covariances are estimated. They also may use different default settings
(than those encountered in MPlus) in terms of the expectation-maximization
algorithm, which can make comparing results across tools challenging.

## The tidyLPA R package

@@

One way that tidyLPA is designed to be easy to use is that it assumes a “tidy”
data structure (Wickham, 2014). This means that it emphasizes the use of a data
frame as both the primary input and output of functions for the package. Because
data is passed to and returned (in amended form, i.e., with the latent profile
probabilities and classes appended to the data) from the function, it makes it
easy to create plots or use results in subsequent analyses. Another noteworthy
feature of tidyLPA is that it provides the same functionality through two
different tools, one that is open-source and available through R, the mclust
package (Scrucca et al., 2017) and one that is available through the commercial
software MPlus (Muthen & Muthen, 1997-2017). Moreover, as both tools use the
same maximum likelihood estimation procedure, they are benchmarked to produce
the same output. Also, note that we have described the model
specifications with descriptions of what is estimated in terms of the variances
and covariances, the common names for the models (i.e., class-varying
unrestricted), and the covariance matrix associated with the parameterizationfor
the six models that are possible to be estimated on the website for tidyLPA (see
here).

## Installation

@@

``` {r, eval = F}
install.packages("tidyLPA")
```

@@

You can also install the development version of tidyLPA, which may have new features that are not yet availalble on CRAN (because the version on CRAN is periodically updated) from GitHub with:

``` {r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("data-edu/tidyLPA")
```

``` {r, message = F}
library(tidyLPA)
library(dplyr) # for preparing the data
```

## Examples

Here, we use a simulated dataset contains data on broad interest, enjoyment, and self-efficacy from the PISA assessment.

First, let's explore the data graphically using the {ggplot2} package:

```{r}
pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>% 
  tidyr::gather(key, val) %>% 
  ggplot(aes(x = val, group = key, fill = key)) + 
  geom_density() + 
  facet_wrap(~key) + 
  theme_bw()
```

Here, we use the {psych} package to explore the distributional properties of the variables:

```{r}
pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>% 
  psych::describe()
```

Add one or two sentences here on the dataset?

### Mclust via tidyLPA

@@

Note that here and throughout this paper we use the function `knitr::kable()` to
format the output in an RMarkdown document, like the one we are using; this is
not necessary to use unless you also wish to format the tidyLPA output in an
RMarkdown document.

``` {r, curry-mac-mclust-example}
pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>%
  single_imputation() %>%
  estimate_profiles(3)
```

### Mplus via tidyLPA

@@

### Fit

We can see a number of statistics for each of the models that were estimated, including the 
AIC (Aikake information criterion; based on -2 * the log-likelihood, and penalized by number of parameters) and BIC (BIC: Bayesian information criterion; based on -2 log-likelihood, and penalized by number of parameters adjusted by sample size). Both are _penalized_ likelihood statistics; the log-likelihood will always decrease with the addition of additional 
paremeters estimated by the model, whereas the AIC and BIC will account for the number of parameters; lower values of
either indicate better fit.

Entropy. `prob_min` refers to the minimum of the diagonal of the average latent class probabilities for most likely class membership, by assigned class. The minimum should be as high as possible, reflecting greater classification certainty (cases are assigned to classes they have a high probability of belonging to; see Jung & Wickrama, 2008); `prob_max`
refers to the maximum. Finally. `n_min` refers to the proportion of the sample assigned to the smallest class (based on most likely class membership); `n_max` to the largest. Finally, `BLRT_p` refers to the p-value for the bootstrapped likelihood ratio test; a value greater than .05 indicates that the model fit better than one with one fewer class. 
Additional fit statistics (other penalized likelihood values, such as the CAIC, SABIC, and ICL), and the bootstrapped LRT 
test statistic, can be obtained with the `get_fit()` function:

```{r, get-fit-m1}
pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>%
  single_imputation() %>%
  estimate_profiles(3) %>% 
  get_fit(m1) %>% 
  knitr::kable()
```

## Comparing a wide range of solutions

@@

````{r, compare-mclust, eval= TRUE}
compare_solutions_results <- pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>%
  single_imputation() %>%
  estimate_profiles(models = c(1, 2, 3, 6),
                    n_profiles = 1:9)

compare_solutions_results

compare_solutions(compare_solutions_results)
```

### Specifying the _model_

We passed the number of profiles, but no details about the model we wished to estimate.

There are a number of different types of models for LPA that estimate different parameters; choosing 
between them is an important consideration, and one that has a bearing on how we will interpret the results we saw above.

Thus, in addition to the number of profiles (specified with the `n_profiles`
argument), the model can be specified in terms of whether and how the variable
variances and covariances are estimated. The models are specified by passing arguments to the `variance` and `covariance`
arguments. The possible values for these arguments are:

@@

In general, the approach to choosing the model is similar to choosing the number of profiles, requiring **deciding on the basis of evidence from multiple sources**, including information criteria, statistical tests, and concerns of interpretability and parsimony. The article by [Pastor and colleagues (2007)](https://www.sciencedirect.com/science/article/pii/S0361476X06000543) has helpful information on the model specifications. Here, the six models that are possible to specify in LPA are described in terms of how the variables used to create the profiles are estimated.

@@

If no values are specified for these, then the variances are constrained to be
equal across classes, and covariances are fixed to 0 (conditional independence
of the indicators).

These arguments allow for four models to be specified:

- Equal variances and covariances fixed to 0 (Model 1)
- Varying variances and covariances fixed to 0 (Model 2)
- Equal variances and equal covariances (Model 3)
- Varying variances and varying covariances (Model 6)

Two additional models (Models 4 and 5) can be fit using MPlus. More information
on the models can be found in the
[vignette](https://data-edu.github.io/tidyLPA/articles/Introduction_to_tidyLPA.html).

## Accessing values in the output

A few helper functions are available to make it easier to work with the output
of an analysis.

`get_data()` returns the data:

``` {r, get-data-for-solution, eval = FALSE}
get_data(compare_solutions_mclust[[32]])
```

We note that `get_data()` returns data in wide format when applied to an object
of class tidyProfile (one element of a tidyLPA object), or when applied to a
tidyLPA object of length one. `get_data()` returns long format when applied to a
tidyLPA object containing multiple tidyProfile analyses (because then the wide
format does not make sense).

To transform data in the wide format into the long format, the `gather()`
function from the **tidyr** package can be used, e.g.:

``` {r, process-data, eval = FALSE}
get_data(compare_solutions_mclust[[32]]) %>% 
  tidyr::gather(Class_prob, Probability, contains("CPROB"))
```

# Recommendations Regarding Analytic Choices

The GRoLTs checklist is a helpful way to think about how best practices for LPA;
here, we show how to implement these using tidyLPA. First, here is the checklist:

```{r, echo = FALSE}
t <- readr::read_csv("GRoLTS-applicability-to-LPA.csv")

tp <- t %>% 
  filter(use == 1) %>% 
  mutate(`LPA Number` = 1:nrow(.)) %>% 
  select(`GRoLTS Number` = Number, `LPA Number`, `Checklist Item`)

### Is a description provided of what variables are related to attrition/missing data?

In LPA, we can determine which parameters - variance
and covariance - are estimated to be the same or different across all profiles,
and, in the case of covariance, whether it is fixed to zero. Of course, for the
profiles we are most interested in, the mean is allowed to vary across profiles.
The models with more parameters freely-estimated use more degrees of freedom;
thus, similar to the number of profiles selected, the balance between adding
more parameters (and having a more complex model) and interpretability and
parsimony must be balanced. If a more complex model fits better and is
interpretable and is justifiable in terms of the data collected, then it may be
preferred to simpler models. If the two models are similar in terms of their
fit, then the more simple, parsimonious model should be selected.

### If classification of cases in a trajectory is the goal, is entropy reported?

# Future Directions and Extensions

First, scholars taking a
person-oriented approach should emphasize reproducibility in carrying out
analyses. This is in part due to the exploratory and interpretative nature of
person-oriented approaches. To this end, presenting multiple models, as in
Pastor et al. (2007), should be encouraged, rather than presenting one solution.
In addition, having multiple analysts review the solutions found is encouraged.
As part of this recommendation, we suggest that researchers consider how their
analyses can be reproduced by other analysts outside of the research team:
Sharing code and data is an important part of this work. Also related,
researchers should consider how profiles are replicated across samples.

A second general recommendation considers more flexibly incorporating time, when
data are collected across multiple time points, into analyses. ISOA groups all
time points and doesn’t make distinctions. Other approaches perform analysis
separately at different timepoints (Corpus & Wormington, 2014). Some integrate
time as a part of the profiles, i.e. growth mixture modeling (groups of
patterns), i.e. within-person growth modeling, where there are individual growth
patterns. Research to date has yet to consider additional challenges in applying
person centered approaches to longitudinal data. For instance, Schmidt et al
(2018) use an Experience Sampling Method (ESM) approach to collecting data and
used a person-centered approach to generate profiles of students in science
class. This work does not account for student-level effects. In other words, it
did not model the shared variance of multiple observations of the same student.

Third, best practices in within-person or longitudinal research call for
modeling the nesting structure. However, researchers have yet to successfully
incorporate this practice into the person centered approach. One way to approach
this is to use cross-classified mixed effects models, as in Strati, Schmidt, and
Maier (2017) and in Rosenberg (2018). In such approaches, dependencies in terms
of, for example, individuals responding to (ESM) surveys at the same time and
repeated responses being associated with the same individuals can both be
modeled, although the effects of these two sources of dependencies are not
nested as in very common uses of multi-level models, but rather are
cross-classified. West, Welch, & Galecki (2014) have a description of the use of
multi-level models with cross-classified data and tools (including those freely
available through R) that can be used to estimate them.

# Implications

# Conclusion

Person-oriented analysis ia way to consider how psychological constructs are
experienced (and can be analyzed) together and at once. Though described in
contrast to a variable-centered approach, scholars have pointed out how
person-oriented approaches are complementary to variable-centered analyses
(Marsh, Ludtke, Trautwein, & Morin, 2009). A person-oriented approach can help
us to consider multiple variables together and at once and in a more dynamic
way, reflected in the methodological approaches for cluster analysis and LPA
that identify profiles of individuals responses.

>>>>

author: 
  - name          : "Joshua Rosenberg"
    affiliation   : "1"
    corresponding : yes
    address       : "1122 Volunteer Blvd., Knoxville, TN, 37996"
    email         : "jmrosenberg@utk.edu"
  - name          : "Caspar van Lissa"
    affiliation   : "2"
  - name          : "Jennifer Schmidt"
    affiliation   : "3"
  - name          : "Patrick Beymer"
    affiliation   : "5"
  - name          : "Daniel Anderson"
    affiliation   : "4"
  - name          : "Matthew Schell"
    affiliation   : "3"

affiliation:
  - id            : "1"
    institution   : "University of Tennessee, Knoxville"
  - id            : "2"
    institution   : "Utrecht University"
  - id            : "3"
    institution   : "Michigan State University"
  - id            : "3"
    institution   : "University of Oregon"
  - id            : "4"
    institution   : "University of Wisconsin, Madison"

abstract: 
keywords          : "Latent Profile Analysis, mixture models, finite mixture models, tutorial, R, MPlus, mclust"
wordcount         : "suggested to be no more than 3,000"

@@

```{r setup, include = FALSE}
library(papaja)
# Seed for random number generation
set.seed(20201229)
library(tidyverse)
library(simstudy)

@@

``` {r, include = FALSE, eval = TRUE}
def <- defData(varname = "x1", formula = 65, variance = 30)
def <- defData(def, varname = "x2", formula = 85, variance = 23)
def <- defData(def, varname = "xMix", 
               formula = "x1 | .5 + x2 | .5", 
               dist = "mixture")
dx <- genData(1000, def)
```

Is there one distribution?

``` {r, echo = FALSE, eval = TRUE}
ggplot(dx, aes(x = xMix)) +
  geom_density() +
  papaja::theme_apa() +
  ylim(0, .10)
```

Or, are there two?

``` {r, echo = FALSE, eval = TRUE}
dx %>% 
  gather(key, val, x1:x2) %>% 
  ggplot(aes(x = val, fill = key)) +
  geom_density(alpha = .3) +
  papaja::theme_apa() +
  ylim(0, .10)
```

This kind of distribution (one that is *bimodal*) is not exclusive to grades.
For teachers, psychologists, researchers, and even friends and family members,
people are highly-complex and not easily able to be characterized by *one*
characteristic or personality trait--and its distribution.

In the social sciences, broadly, and in the psychological sciences, in
particular, a statistical method that can be used to describe how people, in
their individual particularities, may have similarities on the basis of some set
of measures through which they can be grouped in meaningful, distinctive ways.

This approach, which we view as having a provenance in developmental (or
*person-oriented*) approaches (Bergman & El-Khouri, 1997; Magnusson &
Cairns, 1996; see Linnenbrink-Garcia & Wormington, 2017, for a recent review) is
an example of a general mixture model (Harring & Hodis, 2016; Pastor, Barron,
Miller, & Davis, 2007).

In this tutorial, we aim to describe one of the most commonly-used---and
relevant to psychologists---application of the general mixture model, to cases
for which all of the variables for which (relatively) homogeneous groups are
identified from among a (relatively) heterogeneous sample are continuos, *latent
profile analysis* (LPA). After describing the method and some examples of its
use, we provide a tutorial for carrying out LPA in the context of a
freely-available, open-source statistical software package that we have
developed and supported over the past three years for R (R Core Team, 2019),
tidyLPA. Finally, we offer some ideas about best practices and informed
recommendations for researchers aiming to use LPA in their applied work, and
conclude with reflections on the role of statistical software--especially
software that is freely-available, open-source, and highly-performant--in the
psychological sciences.

# Latent Profile Analysis and the Need for Efficient and Reproducible Analyses

The goal of LPA is estimate the parameters for a number of distributions
(typically multivariate) from a single data set. Thus, such an approach is
model-based, and some descriptions in the literature refer to it as model-based
clustering (Hennig, Meila, Murtagh, & Rocci, 2015; Scrucca, Fop, Murphy, &
Raftery, 2017). Thus, one distinction between LPA and other, similar cluster
analytic approaches is that LPA is model-based; instead of using algorithms to
group together cases, LPA seeks to estimate parameters---in terms of variances
and covariances and how they are the same or different across profiles---that
best characterize the different distributions. Then, this approach seeks to
estimate a probability that the observation is a sample from the population
associated with each profile (or mixture component) for each observation in the
dataset.

@@

In general, as more parameters are estimated (i.e., those that are fixed to zero
are estimated as being equal across profiles; or those estimated as being equal
across profiles are freely-estimated across them), the model becomes more
complex; the model may fit better, but also be overfit, meaning that the
profiles identified may be challenging to replicate with another, separate data
set. Even still, flexibility in terms of which models can be estimated also has
affordances. For example, the varying means, equal variances, and covariances
fixed to 0. A researcher might choose this model specification if she wants to
model the variables to be used to create profiles that are independent. This
model is very simple, as no covariances are estimated and the variances are
estimated to be the same across profiles. As we estimate more parameters (and
decrease the degrees of freedom), we are more likely to fit the data, but less
likely to be able to replicate the model with a second set of data. As we
progress toward more complex models (with increasingly complex
parameterization), then we are more likely to fit the data better.
In all, this flexibility associated with LPA also has a more pragmatic cost. It
can be very difficult to efficiently---and reproducibly---estimate and compare a
number of models. This cost means that analysts may focus on a subset of models.
It may also mean that unintentional errors are introduced through copying and
pasting the output of model estimations. Part of why we developed tidyLPA was to
make the model estimation process more efficient and more reproducible. As we
describe later, we do this using R, which has a number of benefits, including
being open-source, powerful (in terms of pre-processing and using the output of
models), and increasingly widely-used in the psychological sciences. We also
provide an interface to perhaps the most widely-used software for LPA, MPlus. In
this way, tidyLPA can both allow researchers to estimate models they already
estimate using MPlus more efficiently and reproducibly, and, for those without
access to this software, to do so entirely using R and the open-source R package
mclust.

# Existing Software Options

There are a number of software options available that provide context for why we 
developed tidyLPA.

@@

In R, a number of tools *can* be used to carry out LPA. OpenMx can be used for
this purpose (and to specify almost any model possible to specify within a
latent variable modeling approach). However, while OpenMx is very flexible, it
can also be challenging to use. Other tools in R allow for estimating Gaussian
mixture models, or models of multivariate Gaussian (or normal) distributions. In
addition to following the same general approach, using tools that are designed
for Gaussian mixture modeling have other benefits, some efficiency-related (see
RMixMod, which uses compiled C++ code) and others in terms of ease-of-use (i.e.,
the plot methods built-in to RMixMod, mclust, and other tools).

However, these existing R package also have some drawbacks in addition to positive features, in that it can be difficult to translate between the model specifications, which are often described in terms of the geometric
properties of the multivariate distributions being estimated (i.e., “spherical,
equal volume”), rather than in terms of whether and how the means, variances,
and covariances are estimated. They also may use different default settings
(than those encountered in MPlus) in terms of the expectation-maximization
algorithm, which can make comparing results across tools challenging.

@@

One way that tidyLPA is designed to be easy to use is that it assumes a “tidy”
data structure (Wickham, 2014). This means that it emphasizes the use of a data
frame as both the primary input and output of functions for the package. Because
data is passed to and returned (in amended form, i.e., with the latent profile
probabilities and classes appended to the data) from the function, it makes it
easy to create plots or use results in subsequent analyses. Another noteworthy
feature of tidyLPA is that it provides the same functionality through two
different tools, one that is open-source and available through R, the mclust
package (Scrucca et al., 2017) and one that is available through the commercial
software MPlus (Muthen & Muthen, 1997-2017). Moreover, as both tools use the
same maximum likelihood estimation procedure, they are benchmarked to produce
the same output. Also, note that we have described the model specifications with
descriptions of what is estimated in terms of the variances and covariances, the
common names for the models (i.e., class-varying unrestricted), and the
covariance matrix associated with the parameterizationfor the six models that
are possible to be estimated.

# A Walkthrough Using the tidyLPA R package

Here, we provide a walkthrough of an analysis using the tidyLPA R package.
## Installation

@@

``` {r, eval = F, echo = TRUE}
install.packages("tidyLPA")
```

@@

You can also install the development version of tidyLPA, which may have new
features that are not yet availalble on CRAN (because the version on CRAN is
periodically updated) from GitHub with:
``` {r gh-installation, eval = FALSE, echo = TRUE}
# install.packages("devtools")
devtools::install_github("data-edu/tidyLPA")
```
``` {r, message = F, echo = TRUE}
library(tidyLPA)
library(dplyr) # for preparing the data
```
## Plotting and Describing the Data

Here, we use a simulated dataset contains data on broad (general, and not situation- or topic-specific) interest, sense of enjoyment, and self-efficacy from the PISA assessment.
First, let's explore the data graphically using the {ggplot2} package:
```{r, echo = TRUE}
pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>% 
  tidyr::gather(key, val) %>% 
  ggplot(aes(x = val, group = key, fill = key)) + 
  geom_density() + 
  facet_wrap(~key) + 
  theme_bw()
```
We see that the data broadly _appears_ to be normally distributed, as the following statistics from the use of the psych package suggest:
``` {r, echo = TRUE}
pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>% 
  psych::describe()
```

## Using tidyLPA directly and exclusively within R

First, we show how to use tidyLPA within R---without any need to use or purchase MPlus.

@@

``` {r, curry-mac-mclust-example, echo = TRUE}
pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>%
  single_imputation() %>%
  estimate_profiles(3)
```

## Using tidyLPA to estimate models efficiently and reproducibly in Mplus

@@

## Determining the Fit of an Estimated Model

We can see a number of statistics for each of the models that were estimated,
including the AIC (Aikake information criterion; based on -2 \* the
log-likelihood, and penalized by number of parameters) and BIC (BIC: Bayesian
information criterion; based on -2 log-likelihood, and penalized by number of
parameters adjusted by sample size). Both are *penalized* likelihood statistics;
the log-likelihood will always decrease with the addition of additional
paremeters estimated by the model, whereas the AIC and BIC will account for the
number of parameters; lower values of either indicate better fit.

Entropy. `prob_min` refers to the minimum of the diagonal of the average latent
class probabilities for most likely class membership, by assigned class. The
minimum should be as high as possible, reflecting greater classification
certainty (cases are assigned to classes they have a high probability of
belonging to; see Jung & Wickrama, 2008); `prob_max` refers to the maximum.
Finally. `n_min` refers to the proportion of the sample assigned to the smallest
class (based on most likely class membership); `n_max` to the largest. Finally,
`BLRT_p` refers to the p-value for the bootstrapped likelihood ratio test; a
value greater than .05 indicates that the model fit better than one with one
fewer class. Additional fit statistics (other penalized likelihood values, such
as the CAIC, SABIC, and ICL), and the bootstrapped LRT test statistic, can be
obtained with the `get_fit()` function (processed with the tidyr R package to be in "long" rather than "wide" format):

``` {r, get-fit-m1, echo = TRUE}
library(tidyr)

pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>%
  single_imputation() %>%
  estimate_profiles(3) %>% 
  get_fit(m1) %>% 
  gather(stat, val)
```

## Comparing the Fit for Many Estimated Models

@@

````{r, compare-mclust, eval = TRUE, echo = TRUE}
many_analyses_1_3 <- pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>%
  single_imputation() %>%
  estimate_profiles(models = c(1, 3),
                    n_profiles = 1:5)

compare_solutions(many_analyses_1_3)
```

## Inspecting and Plotting the best-fitting estimated models

It appears that the four and five-class solutions fit best. Let's inspect those, specifically:

```{r, echo = TRUE, eval = TRUE}
models_4_5 <- pisaUSA15[1:100, ] %>%
  select(broad_interest, enjoyment, self_efficacy) %>%
  single_imputation() %>%
  estimate_profiles(models = 1,
                    n_profiles = c(4, 5))

models_4_5 %>% 
  plot_profiles(add_line = TRUE)
```

Here, it appears that the addition of a fifth class does not add a great deal relative to the four-class solution. Accordingly, we consider the four-class solution to be the solution we interpret and present, while heeding the recommendations about reporting which other models we estimated, how we chose to focus on the two best-fitting models we did, and why we presented the four-class, class-invariant set of estimates.

Were we to wish to inspect _only_ that chosen model, we could estimate the model again, replacing `n_profiles = c(4, 5)` with `n_profiles = 4`. Alternatively, we could use list-indexing within R to index only that model. Note that the object `models_4_5` below includes the four and five-class estimated models, in that order. Using `[[1]]` indexes---selects---only the first of the two models, whereas `[[2]]` indexes the second. We index the four-class solution using `[[1]]`, and save it to a new object.

```{r}
model_4 <- models_4_5[[1]]
```

Using this single estimated model, we can also create a faceted plot of density plots for an object of class 'tidyLPA'. For each variable, a Total density plot will be shown, where cases are weighted by the posterior probability of being assigned to that class:

```{r}
model_4 %>% 
  plot_profiles(add_line = TRUE)
```

## Visualizing and Examining a Chosen Solution Further
Last, we can create a faceted plot of two-dimensional correlation plots and unidimensional density plots for a single set of model estimates:
```{r}
model_4 %>% 
  plot_bivariate()
```
We can use `get_data()` to return data in wide format when applied to an object
of class tidyProfile (one element of a tidyLPA object)

`get_data()` returns long format when applied to a
tidyLPA object containing multiple tidyProfile analyses (because then the wide
format does not make sense).

To transform data in the wide format into the long format, the `gather()`
function from the **tidyr** package can be used:

``` {r, process-data, eval = TRUE, echo = TRUE}
get_data(model_4) %>% 
  tidyr::gather(Class_prob, Probability, contains("CPROB"))
```

In addition to the pragmatic challenges facing the analyst carrying out LPA, there 
are a number of statistical issues that anaylsts face. Past research is informative in
this respect; Bauer

## Recommendations Regarding Analytic Choices

The GRoLTs checklist, developed for a particular use of the general mixture
model---a growth mixture model---is a helpful way to think about how best
practices for LPA [@van2017grolts]. We interpret them in light of LPA; note that
the full checklist is not applicable (as, for example, guidelines about
trajectories over time are not pertinent to LPA), and we only present those that
are.

First, here is the checklist:

``` {r, echo = FALSE}
t <- readr::read_csv("GRoLTS-applicability-to-LPA.csv")

tp <- t %>% filter(use == 1) %>% mutate(`LPA Number` = 1:nrow(.)) %>%
  select( `LPA Number`, `GRoLTS Number` = Number, `Checklist Item`)

As missing data can have a bearing on the estimates, an important consideration is whether or not the missing data is associated with the levels of the variables---or not. Consider consulting general guidelines on missing data, including @enders2010applied, @little2019statistical, and @allison2001missing. 
### Is a description provided of what variables are related to attrition/missing data?

Note whether any demographic or other variables are related to missing data.

Because LPA cannot accomodate any missing values (and so uses listwise deletion) for this recommendation, tidyLPA has a function for addressing missing data that are based upon single imputation (with two methods available, one using a contemporary, machine learning-based procedure).

The variables can be plotted and described prior to the analyses. While not built-in, we show later how this can be done using the ggplot2 R package, and the `describe()` function from the psych R package.

The name and version of the software can be mentioned. Within R, a citation including the version can be generated for any R package, including tidyLPA, by using the `citation()` function (e.g., `citation("tidyLPA")).

In LPA, we can determine which parameters - variance and covariance - are
estimated to be the same or different across all profiles, and, in the case of
covariance, whether it is fixed to zero. Of course, for the profiles we are most
interested in, the mean is allowed to vary across profiles.

In general, the approach to choosing the model is similar to choosing the number
of profiles, requiring **deciding on the basis of evidence from multiple
sources**, including information criteria, statistical tests, and concerns of
interpretability and parsimony. The article by [Pastor and colleagues
(2007)](https://www.sciencedirect.com/science/article/pii/S0361476X06000543) has
helpful information on the model specifications.

Here, the six models that are possible to specify in LPA are described in terms of how the variables used to create the profiles are estimated. In tidyLPA, the models are specified by passing
arguments to the `variance` and `covariance` arguments. The possible values for
these arguments are:

@@

These models are described in Appendix A.

MPlus uses random starts to initialize the estimation, whereas mclust (the R package) does not. Thus, for users of MPlus, the number of random starts and final iterations should be reported. By default, tidyLPA uses the MPlus defaults. Information on the number of random starts can be found in the model object returned by the tidyLPA function `estimate_profiles()`. Different random starts can be passed by passing arguments to the `ANALYSIS` code for MPlus, e.g.:

```{r, eval = FALSE}
estimate_profiles(3, package = "Mplus", 
                  ANALYSIS = "starts = 100, 20;")
```

We provide a number of functions for model comparison. The `compare_solutions()` profile
takes as an argument a particular measure of fit (defaulting to the BIC), and returns information
for all of the models estimated. It also implements an ensemble, analytic hierarchiy approach, that uses weighted values of a number of fit indices; see [@akogul2017approach] for more.

The number of fitted models can easily be reported based on the tidyLPA code run. For example, for the following code, models with 1-6 profiles, and the first (class invariant) and third (class invariant, unrestricted) model parameterizations are estimated:

```{r, eval = FALSE}
estimate_profiles(n_profiles = 1:6, 
                  models = c(1, 3))
```

Determining the number of cases per class can easily be done by using the `get_data()` function from tidyLPA, passing the estimated model object from tidyLPA.; see the `Class` column (as well as the posterior probability estimates for every class for every observation).

This is a recommendation which can be difficult to achieve using existing software. tidyLPA provides two functions, `plot_profiles()` and `plot_bivariate()`, to plot the estimates.

The above-mentioned tidyLPA functions also plot the raw data with the estimates (by default).

These characteristics can be obtained by using the `get_estimates()` function, passing the estimated model object from tidyLPA.

The R files used for the analyses can easily be included in online repositories and supplementary online materials (as `.R` or `.Rmd` files).

# Summary and Conclusion

LPA can be used to evaluate how psychological constructs are experienced (and
can be analyzed) together and at once. Though described in contrast to a
variable-centered approach, scholars have pointed out how person-oriented
approaches, including LPA, are complementary to variable-centered analyses
(Marsh, Ludtke, Trautwein, & Morin, 2009). Our aim was to describe how
researchers may get started in an informed way as researchers seek to understand
how individuals interact, behave, and learn in ways that embraces the complexity
of these experiences.

# Appendix A

@@
